Learning how to act: making good decisions with machine learning

### 2017-07-20 thesis/offcuts.tex
starts on chapter 1
working on intro

@@

>>>>

\begin{itemize}
\item Speech recognition (for systems like Siri or Google)
\item Machine translation 
\item Image classification
\item Forecasting the weather
\item Playing Go 
\item Identifying spam emails
\item Automated essay marking
\item Predicting the risk of death in patients with pneumonia.
\item Predicting who will re-offend on release from prison 
\item Predicting which customers will cease to be your customers
\item Demand prediction for inventory control
\item Predicting who will click on an ad
\item Financial trading
\item Recommending movies
\item Online search
\item Self driving cars
\item Pricing insurance
\end{itemize}

@@

connection to bias variance trade-off. Increasing the assumptions we impose via a model introduces bias but decreases variance in the result.

For example, in Australia we might detect that driving on unpaved roads is associated with poorer health outcomes. It is plausible that there are some negative health impacts of bumpy roads, but more likely the correlation occurs because unpaved roads and poor access to health services are both a \emph{consequence} of living in regional areas.

Problems highlighted as intractable in the reverse causal sense are also intractable in the forward inference form, typically because concern situations for which we do not have a sufficient number of similar instances to allow statistical reasoning. For example, the war question posed in Gelman.

Observing a system, without the ability to intervene or experiment on it. Fundamental challenge because larger samples are not enough.

In this case, the observed correlation is not an artifact of random noise in a sample of data. We would expect the association to continue to hold if sampled new data.

### 2017-07-20 thesis/thesis_map.tex
working on intro

\section{What is causality and why do we care? }

The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still widely viewed as poorly defined. This has led to a reluctance among applied researchers in many fields to refer to causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}. However, the magnitude, direction and even existence of an association depends on what other variables we adjust for (or include in a regression). Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via common sense the \emph{implications} of the reported associations.

We discuss more detailed definitions of causality in section c

The what-if type questions from the why. \citep{Gelman2010} \citep{}. Why do whites to better than blacks in school (in America). Suggests that reverse causal inference questions are more interesting and motivate most of social science.

I do not find this distinction useful. 
We can only change the future - history is useful only as far as it tells us something about the future.
Reverse causal questions can be reposed as forward ones, when making the translation a reverse causal query will be effectively asking about many possible interventions (rather than just one). Problems highlighted as intractable in the reverse causal sense are also intractable in the forward inference form, typically because concern situations for which we do not have a sufficient number of similar instances to allow statistical reasoning. For example, the war question posed in Gelman.

A distinction between forward causal inference, what happens if we do X and reverse casual inference

To explain or to predict \citep{Shmueli2010}
The two cultures \citep{Breiman2001}

\paragraph*{There are two reasons why correlation is not causation}. The first is related to variance and over-fitting. Observations are noisy. With a finite data set with enough variables we will be able to find some that are completely unrelated but correlate purely by chance. The second arises from bias, typically introduced by an un-observed confounding variable. In this case, variables are correlated not by chance. We would expect the relationship to hold if we sampled more data. However, they are not causally related in that intervening to set one would not likely effect the other. EXAMPLE WITH FIGURE.

\subsection{Defining causality}
\begin{itemize}
\item what is explanation?
\item any model that aims to predict the outcome of an action or intervention in a system
\item I do not see the distinction between explanation and (causal) prediction. Explanation is all about the ability to compress and to generalise. The more a model can do this, the more we view as providing an understanding of the why. 
\item mediation?
\end{itemize}

\section{What problems require causal inference?}

\subsubsection{Examples of typical machine learning problems. Are they causal?}

Consider the following problems, which span a wide range of the types of questions machine learning is currently being applied to. Which of them require casual inference? How can we identify characteristics of a problem that make causal modelling important?

@@

Asking if these problem require causal inference is somewhat disingenuous as they are not posed in sufficient detail to determine if causality is an important consideration. In particular, I failed to specify how any (machine learning) model we might build would be used, and what actions could be taken in response to its predictions. This is a trivial but significant observation.

It comes down to asking, is the objective or loss function we are minimising a good proxy for our underlying goal in building the model.

@@

What about predicting the risk of death in patients with pneumonia? Suppose the goal is to build a model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective for serious cases, this seems like a straightforward prediction problem. It is not. Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if the model is used to make admition decisons. \citet{Caruana2015} found exactly this effect in a real data set. The model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pneumonia. They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk. The issue is not with model, it performed very well at the task for which it was trained, which is to predict wo would be likely to die under the original admition and treatment protocols. However, using it to decide how to \emph{change} these protocols could kill. The actual question of interest in this case is what happens to patients with charachteristics $X$ if we assign them treatment according to decision rule $Z$.

Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc). For all these problems the assumptions required to treat them as pure prediction problems are 
\begin{enumerate}
\item The treatment is assumed to be effective (at least better than nothing)
\item Deciding who to treat on the based of the model predictions won't change the relationship between features and outcome 
\end{enumerate}

@@

Playing Go (and other games) is a case with some subtleties. At every turn, the AI agent has a number of actions available. The board state following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning based on millions of previous games to estimate the probability that each of these board states will lead to a win. ... this is interesting maybe come back to it ... an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ... One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken). When can we learn from the actions others have taken? When there is no confounding. And does this hold with Go? Probably because the board state encapsulates everything that should determine what move is played. Learning directly from actions and trying to generalise (can in some instances reduce the problem to standard ML)

@@

\begin{itemize}
\item Does acting on the predictions of the model change the mapping from features to target? (at least if the decision process is open to scrutiny). In general, if we believe that humans are generally trying to optimise to various goals of their own then for any system interacting with them the answer to this will be yes.
\item

@@

@@

@@

\section{Approaches to causality}

@@

@@

\section{Causal Discovery}

@@

>>>>

\section{What is causality? }

The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still widely viewed as poorly defined. This has led to a reluctance among applied researchers in many fields to refer to causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}. However, the magnitude, direction and even existence of an association depends on what other variables we adjust for (or include in a regression). Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via \quotes{common sense} the implications of the reported associations.

There are two ways in which an association detected in a dataset may be non-causal. The first is that the variables concerned may not be related at all, and the association has arisen by chance in that data sample. Given a finite data on enough variables, there is a high probability of finding some that appear correlated even though they are completely unrelated. For example, based on data from the years 1999 to 2009, the age of miss America is strongly correlated with the number of murders  (in the US) by steam, hot vapours and hot objects \citep{Vigen2015}. We would not expect this relationship to hold if we were to look at a new sample of data. This form of spurious correlation also has serious repercussions. It lies at the heart of major problems with the process of scientific research: researchers are incentivised to detect effects and thus to explore many possible paths in the process of analysing data and studies that fail to find an effect are less likely to be published. As a consequnce, the likelyhood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}. This issue is also highlighted by recent crises in replication \citep{OpenScienceCollaboration2015}. This issue can be ameliorated by getting more data and by seperating learning models from evaluating their performance, for example by evaluating models on a stict hold-out set or on the extent to which their results can be replicated.

However, a strong association, observed in multiple indpendent studies may still not be causal. The correlation can arise because both variables are consequences of some other, unmeasured factor. For example, the reading ability of children under twelve is strongy correlated with their height, because older children are taller and can read better. However height is not a cause of reading ability because if we were to intervene to increase a child's height, for example by giving them growth hormones, we would not expect dramatic improvements in their reading. Similarly, extra tutoring in reading will not make a child grow taller. This problem is fundamentally different to the issue of spurious correlations arising by chance in finite datasets. Obtaining more (even infinitely many more) samples without directly intervening in the system to manipulate the variables does not allow us to seperate causation from correlation.

The key distinction between a real, but non-causal, association and a causal relationship is in what happens if we intervene in the system and change one of the variables. In this thesis, I take an interventionalist viewpoint of causality: any model or approach designed to predict the outcome of intervening in a system is causal. This viewpoint captures the types of questions that motivate this thesis. How can we change the way we do things to obtain better outcomes.

Causality is often liked to explaination; understanding how and why things happen. I view explaination in terms of compression and generalisabilty: how much information about the world a model can capture. This creates a hierachy in the degree to which models are explainitory, rather than a simple binary distinction. A standard predictive model encodes all the information we need to predict some output given inputs provided the system generating the data does not change. A high-level causal model might be able to predict the outcome of a specific intervention holding all else fixed. More detailed causal models could predict the outcome for a wide range of combiations of interventions conditional on a range of contexts. By considering conditional interventions within our definition of causal questions we also capture mediation: the study of through which pathways one variable causes another \cite{vanderweele2015explanation}. Finally, a model that can distil how elements interact to mathematical equatons like Newton's laws can be used to predict what will happen in an astounding range of settings, including many we have never previously observed \footnote{Although Newton's laws are not fully general, as they break down for extremely massive objects}.

\citet{Gelman2010,gelman2013ask} make a distinction between forward causal inference, the types of \quotes{What if} questions I focus on in this thesis, and reverse causal questions, asking why something occurs. The former aims to identify the effect of a known cause. The latter can be viewed as identiying causes of an effect. They regard forward causal inference as well defined within the counterfactual and graphical model frameworks for causal inference, which we describe in chapter \ref{chap:causal_models}. However, state that \quotes{A reverse causal question does not in general have a well-defined answer, even in a setting where all possible data are made available.} I view this as overly pessimistic, depending on how \quotes{all possible data} is defined. The goal of identifying the causes of an effect can be formalised within the graphical causal model framework. Solving this problem is certainly much more challanging than identifying the effect of a specific intervention on given outcome, since it requires us to test or infer the effect of interventions on many different variables. These practical difficulties may well be overwealming, particularly in fields such as social science and economics where datasets are relatively small, systems are complex, the variables are difficult to directly manipulate and even relatively simple "What if" questions are hard to resolve conclusively. However, this does not mean that the problem of identifying causes of effects is ill-posed in principle. It can be viewed as a form of causal discovery: the attempt to learn the structure of the causal relationships between variables, on which there is a rich liturature which we review briefly in section \ref{sec:causal-discovery}.

There has traditionally been a large gap between researches in machine learning who focus on prediction, use largly non-interpretable models and researches in statistics, social science and economics who (at least implicitly) aim to answer causal questions and tend to use heavily theory driven models. However, there is relatively little awareness, particularly within the machine learning and data science communities, of what constitutes a causal problem and the implications of for the training and evaluation models. In the next section we emphasise the subtley that can exist in determining if a problem is causal by examiming some typical examples.

\section{What makes a causal inference problem?}

Machine learning is in the midst of a boom, driven by the availability of large datasets and the computation resources to process them. Machine learning techniques are being applied a huge range of problems, in both industry and achademia. The following examles are inteded to capture the bredth of problems that machine learning algorithms are actively beeing applied to. Which, if any, of these problems require causal inference?

@@

The question is disingenuous because I have not posed the problems in sufficient detail to determine if causality is an important considertion. In particular, I failed to specify how any model we might build would be used: what actions would be taken in response to its predictions.

@@

What about predicting the risk of death in patients with pneumonia? Suppose the goal is to build a model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective for serious cases, this seems like a straightforward prediction problem. It is not. Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if the model is used to make admition decisons. \citet{Caruana2015} found exactly this effect in a real data set. The model learnt that people suffering asthma were \emph{less} likely to die from pneumonia. They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk. The issue is not with model, it performed very well at the task for which it was trained, which is to predict wo would be likely to die under the original admition and treatment protocols. However, using it to decide how to \emph{change} these protocols could kill. The actual question of interest in this case is what happens to patients with charachteristics $X$ if we assign them treatment according to decision rule $Z$.

Predicting which customers will leave or who will re-offend if granted parole also fit within the category of problems where you wish to identify a group for which a problem will occur and target some treatment to them (hospitalisation, loyalty reward, more monitoring \& support on parole, etc). Predictive models can be applied to such problems provided it is known which treatment option will be most effective for the target group and deciding who to treat on the based of the model predictions won't change the relationship between features and outcome.

@@

Playing Go (and other games) is a case with some subtleties. At every turn, the AI agent has a number of actions available. The board state following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning, based on millions of previous games, to estimate the probability that each of these reachable board states will lead to a win \footnote{This is a challenging pattern recognition problem. There are around $2 \times 10^{170}$ legal board positions in Go, (\citep{tromp2016number}), so the algorithm cannot simply memorise the proportion of times each state leads to a win, it must identify higher level features of the board state that are associated with winning.}. Supervised learning can also be applied to learn a policy $\P{a|s}$, the probability of selecting action $a$ given board state $s$. This allows the agent to estimate the likelyhood of winning from a given starting state by simulating (many times) the remainder of the game, drawing actions from $\P{a|s}$ for both players. Google's Alpha Go, which in May 2017 beat the then strongest human player \citep{Mozur2017}, incorporates a compbination of these approaches \citep{silver2016mastering}. The supervised learning was enhanced by having the agent play (varients) of itself many times, so that its estimates of value for each board state and of the likelyhood an opponent will play as given move are based on a combination of replicating the way humans play and on the moves that led to a win when playing itself.

The problem of playing go is clearly causal from the interventionalist perspective. The agent wishes to learn the probability of a win given an action they take. However,

... this is interesting maybe come back to it ... an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ... One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken).

When can we learn from the actions others have taken? When there is no confounding. And does this hold with Go? Probably because the board state encapsulates everything that should determine what move is played. Learning directly from actions and trying to generalise (can in some instances reduce the problem to standard ML)

@@

This is a trivial but significant observation. It comes down to asking, is the objective or loss function we are minimising a good proxy for our underlying goal in building the model. 
\begin{itemize}
\item Does acting on the predictions of the model change the mapping from features to target? (at least if the decision process is open to scrutiny). In general, if we believe that humans are generally trying to optimise to various goals of their own then for any system interacting with them the answer to this will be yes.
\item

@@

How can we characterise which problems are predictive and which are causal?

@@

Always, unless you can demonstrate otherwise.

@@

\section{Approaches to causality}
This stuff gets built into explaining how the thesis fits together.

@@

\section{Granger causality}

Mentioned in the category of causal frameworks,

Granger causality is less of a framework for defining causality and on which a wide range of algorithms and methods can be constructed, but a method for infering causal structure from data. As such is discussion really belongs in section \ref{sec:causal-discovery}.

TODO move this to a paragraph at the end of the comparing and unifying section. \todo{}

@@

\section{Causal Discovery}
\label{sec:causal-discovery}

@@

\chapter{Causation vs Prediction}

The two cultures \citep{Breiman2001}
 - wrong because suggets we can just use the machine learning approach for everything. Does not address the problems that a) structural assumptions required for causal inference. b) we cannot use cross-vaidation to assess the validity of a causal model (really b is the whole point).

To explain or to predict \citep{Shmueli2010}
- wrong because they assume all the parts of explanatory modelling are required - so black box models cannot be used.

Predict the relationship between X and Y as a way to stratify to reduce the variance in an experiment.

They consider a different definition of explanatory models,

Definition as backwards. Explanatory models focus on theory because some theory (or equivelently stated assumptions) are required in order to predict the outcome of an intervention from observational data.

Causal modelling frameworks allow us to seperate assumptions about the structure of the relationships between variables from assumptions about the functional form.

What can be data driven and what cannot? (for a given set of assumptions).

Allowing for the possibility that we might only need to be able to predict the outcome of an intervention on a given subset of variables (for example those over which we have the ability to manipulate) or the possibility that we can can obtain observation from directly intervening in a system creates creates a need for models that are neither explanatory or predictive in the languate of EoP but are built on a hybrid of both approaches.
